Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-20 paper-amia-summit-2018/amia_summit_2018.tex
update RNN figure
fixed spelling mistakes

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. 
%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. 
Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.  
%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information is learned from the dataset, and used as a ``model'' to classify sequences.

@@

One of the model-based classification methods is probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM). Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step. Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory. It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}. A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}. In Lipton et. al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc. LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit. A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed. These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.

@@

Our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level. In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity data set. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. That RNN can capture long-term dependencies for predicting the future events is its main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence. Then, we applied a softmax layer at each time-step for predicting the target which is the label of the entire sequence (successful or unsuccessful). This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training data set by manually annotating the data set using the MYSCOPE codebook. 
%\pagebreak

>>>>

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. 
%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. 
Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.  
%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information is learned from the dataset and used as a ``model'' to classify sequences.

@@

One of the model-based classification methods is a probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM). Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step. Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory. It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}. A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}. In Lipton et. al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc. LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit. A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed. These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.

@@

Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level. In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity dataset. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. That RNN can capture long-term dependencies for predicting the future events is its main advantage over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn_small.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence. Then, we applied a softmax layer at each time-step for predicting the target which is the label of the entire sequence (successful or unsuccessful). This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset by manually annotating the dataset using the MYSCOPE codebook. 
%\pagebreak
